## Exemplar: Standardized effect size {#effectsize_exemplar_standardized}

<mark>
This section is in *alpha*. We welcome help and feedback at all levels!
If you would like to contribute, please see
[Contributing to the Guidelines](https://github.com/transparentstats/guidelines/wiki/Contributing-to-the-Guidelines).
</mark>

```
TODO: This needs a domain where we can argue that Cohen's d is an exemplar analysis, then repeat structure of exemplar 1 with it
May be an example of existing meta-analysis in HCI.
```

### Libraries needed for this analysis

```{r es-standard-setup, warning = FALSE, message = FALSE}
library(tidyverse)
library(effsize)  # for cohen.d()
```

```{r es-simple-boilerplate, include = FALSE}
format_num <- function(nums, sigdigits = 3) gsub("\\.$", "", formatC(nums, sigdigits, format = "fg", flag="#"))
```


### Standardized effect size

```{r es-standard-data_generation}
set.seed(12)
n <- 20
data <- tibble(
  group = rep(c("A", "B"), each = n),
  completion_time_ms = c(
    rlnorm(n, meanlog = log(170), sdlog = 0.3),
    rlnorm(n, meanlog = log(50), sdlog = 0.4)
  )
)
```

```{r}
cohen_d <- cohen.d(completion_time_ms ~ group, data = data)

# manual calculation
data_A <- (data %>% filter(group == "A"))[["completion_time_ms"]]
data_B <- (data %>% filter(group == "B"))[["completion_time_ms"]]
sd_A <- sd(data_A)
sd_B <- sd(data_B)
sd_pool <- sqrt( (sd_A^2 + sd_B^2) / 2 )
cohen_d_manual <- abs(mean(data_A) - mean(data_B))/sd_pool
```

**Standardized effect size:** Cohen's d = `r format_num(cohen_d$estimate)` SDs with 95% confidence interval [`r format_num(cohen_d$conf.int[1])` , `r format_num(cohen_d$conf.int[2])`]


