## Exemplar: Nonparametric effect size {#effectsize_exemplar_nonparametric}

For a nonparametric test that produces a Z-score, like the Mann-Whitney U test or the Wilcoxon Signed-Rank test, an effect size can be computed as:

<center>$r = \left|\frac{Z}{\sqrt{N}}\right|$</center>

Above, Z is the Z-score and N is the number of observations in all groups [@Rosenthal1991a, p. 19). The result, *r*, is a variance-based effect size, like Pearson *r*, not a Cohen *d*-family effect size. The *r* can be squared to estimate the percentage of variance explained, however it will not be exactly equivalent to the Pearson *r*. 

```
TODO: This needs a domain where we can argue that the nonparametric approach is an exemplar analysis, then repeat structure of exemplar 1 with it
```

### Libraries needed for this analysis

```{r es-nonparametric-setup, warning = FALSE, message = FALSE}
library(tidyverse)
library(coin)        # for wilcox_test
```

```{r nonparametric-boilerplate, include = FALSE}
format_num <- function(nums, sigdigits = 3) gsub("\\.$", "", formatC(nums, sigdigits, format = "fg", flag="#"))
```


### Nonparametric effect size

```{r es-nonparametric-data_generation}
set.seed(12)
n <- 20
data <- tibble(
  group = rep(c("A", "B"), each = n),
  completion_time_ms = c(
    rlnorm(n, meanlog = log(170), sdlog = 0.3),
    rlnorm(n, meanlog = log(50), sdlog = 0.4)
  )
)
```

```{r}
data_A <- (data %>% filter(group == "A"))[["completion_time_ms"]]
data_B <- (data %>% filter(group == "B"))[["completion_time_ms"]]
wilcox_result <- wilcox_test(completion_time_ms ~ factor(group), data = data)
effect_r <- abs(wilcox_result@statistic@teststatistic / sqrt(nrow(data)))
```

**Non-parametric effect size:** Variance-based effect size *r*  = `r format_num(effect_r, 2)`.
